Cell type assignments for SCPCP000015

### 2025-02-03 analyses/cell-type-ewings/exploratory_analysis/08-merged-celltypes.Rmd
initial addition of notebook annotating merged object

```{r}
all_results_df <- prep_results(
  sce, 
  singler_df = singler_df, 
  cluster_df = NULL, 
  aucell_df = aucell_df,
  cluster_nn = params$cluster_nn,
  cluster_res = params$cluster_res,
  join_columns = c("barcodes", "library_id")
  ) |>
  dplyr::mutate(barcodes = glue::glue("{library_id}-{barcodes}"))

@@

### Cluster assignments

The below UMAP shows the cluster assignments output by `evaluate-clusters.sh` using the following parameters:

- Leiden with modularity 
- Nearest neighbors: `r params$cluster_nn`
- Resolution: `r params$cluster_res`

```{r, fig.height = 5}
plot_faceted_umap(all_info_df, cluster, legend_title = "Cluster")
```

@@

```{r, fig.height=10}
auc_columns <- colnames(all_info_df)[which(startsWith(colnames(all_info_df), "auc_"))]
cluster_density_plot(all_info_df, auc_columns, "AUC")
```

@@

```{r}
all_info_df <- all_info_df |> 
  dplyr::arrange(singler_lumped)
head(all_info_df)

annotated_exp_heatmap(
  all_info_df, 
  exp_columns = auc_columns, 
  cell_type_column = "singler_lumped", 
  cluster_column = "singler_lumped", 
  legend_title = "AUC"
)
```

@@

```{r, fig.height=8}
ggplot(exp_df, aes(x = mean, color = tumor_cell_classification, fill = tumor_cell_classification)) +
  geom_density(alpha = 0.5, bw = 0.1) +
  facet_wrap(vars(geneset)) +
  theme(
    aspect.ratio = 1,
    strip.background = element_rect(fill = "transparent", linewidth = 0.5),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
  )
```

@@

```{r, fig.height=7}
mean_exp_columns <- colnames(all_info_df)[which(endsWith(colnames(all_info_df), "_mean"))]
cluster_density_plot(all_info_df, mean_exp_columns, "mean gene expression")
```

>>>>

```{r}
all_results_df <- prep_results(
  sce, 
  singler_df = singler_df, 
  cluster_df = NULL, 
  aucell_df = aucell_df,
  consensus_df = consensus_df,
  cluster_nn = params$cluster_nn,
  cluster_res = params$cluster_res,
  join_columns = c("barcodes", "library_id")
  ) |>
  dplyr::mutate(barcodes = glue::glue("{library_id}-{barcodes}"))

@@

### Consensus cell type assignments

```{r, fig.height=5}
plot_faceted_umap(all_info_df, consensus_lumped, legend_title = "Consensus cell types") +
  theme(strip.text = element_text(size = 8))
```

@@

```{r, fig.height=10}
auc_columns <- colnames(all_info_df)[which(startsWith(colnames(all_info_df), "auc_"))]
cluster_density_plot(all_info_df, auc_columns, "consensus_lumped", "AUC")
```

Here we should identify any tumor cells as those that have high expression of `EWS-FLI` targets and low expression of `EWS-FLI` repressed targets. 
I think we want to use the gene sets where we see some separation in AUC values between cell types to do our assignment.

- `aynaud-ews-targets`: > 0.05
- `staege`: > 0.25
- `riggi_down`: < 0.025
- `migayawa_down`: < 0.03

Intriguingly we see high expression of `hallmark_EMT` and the down gene sets (genes that are repressed by `EWS-FLI`) in fibroblasts, which also happen to have high expression of the `EWS-FLI` low targets identifed by Wrenn et al.
That paper identifies `EWS-FLI` low cells as cancer associated fibroblasts.
Let's label these cells accordingly. 
If cells have `in_geneset=TRUE` for the `hallmark_EMT`, `wrenn-nt5e-genes`, and `migayawa_down` they will be classified as `ews_low_cafs`.

- `wrenn-nt5e-genes`: > 0.1
- `hallmark_EMT`: > 0.05
- `migayawa_down`: > 0.03
- `riggi_down`: > 0.025

```{r}
all_info_df <- all_info_df |> 
  dplyr::mutate(
    final_annotation = dplyr::case_when(
      `auc_wrenn-nt5e-genes` > 0.1 & auc_miyagawa_down > 0.03 ~ "ews_low_caf",
      `auc_aynaud-ews-targets` > 0.025 | 
        consensus_annotation %in% c("Unknown", "muscle cell", "smooth muscle cell") ~ "ews_high",
      .default = consensus_annotation
    )
  )

head(all_info_df)
table(all_info_df$final_annotation)
```

@@

```{r}
all_info_df <- all_info_df |> 
  dplyr::arrange(final_annotation)

annotated_exp_heatmap(
  all_info_df, 
  exp_columns = auc_columns, 
  cell_type_column = "consensus_lumped", 
  cluster_column = "final_annotation", 
  legend_title = "AUC"
)

# get annotation colors for cell type and cluster 
cell_types <- unique(all_info_df$final_annotation)
num_cell_types <- length(cell_types)
cell_type_colors <- palette.colors(palette = "alphabet") |>
  head(n = num_cell_types) |>
  purrr::set_names(cell_types)

# create annotation for heatmap
annotation <- ComplexHeatmap::columnAnnotation(
  cell_type = all_info_df$final_annotation,
  col = list(
    cell_type = cell_type_colors
  )
)

# build matrix for heatmap cells x gene set/ expression columns 
heatmap_mtx <- all_info_df |>
  dplyr::select(barcodes, all_of(auc_columns)) |>
  tibble::column_to_rownames("barcodes") |>
  as.matrix() |>
  t()

# remove any extra strings from the expression column names 
rownames(heatmap_mtx) <- stringr::str_remove(rownames(heatmap_mtx), "auc_|_mean")

# plot heatmap of marker genes
plot_gene_heatmap(heatmap_mtx,
                  row_title = "",
                  legend_title = "final_annotation",
                  annotation = annotation,
                  cluster_columns = TRUE
)
```

@@

```{r, fig.height=8}
ggplot(exp_df, aes(x = mean, color = tumor_cell_classification, fill = tumor_cell_classification)) +
  geom_density(alpha = 0.5, bw = 0.5) +
  facet_wrap(vars(geneset)) +
  theme(
    aspect.ratio = 1,
    strip.background = element_rect(fill = "transparent", linewidth = 0.5),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
  )
```

@@

```{r, fig.height=30, warning=FALSE}
mean_exp_columns <- colnames(all_info_df)[which(endsWith(colnames(all_info_df), "_mean"))]
cluster_density_plot(all_info_df, mean_exp_columns, annotation_column = "consensus_annotation", "mean gene expression")
```
